Forecasting Influenza Hospitalizations Using a Bayesian Hierarchical Nonlinear Model with Discrepancy

### 2024-12-18 nonlinear_forecast/manuscript.tex
update for submissions

% \documentclass[ba,preprint]{imsart}% use this for supplement article
\documentclass[ba]{imsart}

%% Packages
\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
\RequirePackage[numbers]{natbib}
%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue,backref=page,backref=page]{hyperref}
\RequirePackage{graphicx}

%%%%%%%%%%%packages added by Spencer%%%%%%%%%%%%%
\usepackage{booktabs} %for the table in the analysis section
\usepackage{enumerate} %for the short algorithm
\usepackage{subfigure} %for making plots with multiple images
\usepackage{tabularx} %for the spacing in tabular
\usepackage{newtxmath} %for making indicator function
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\pubyear{2024}
\arxiv{2010.00000}
\volume{TBA}
\issue{TBA}
\firstpage{1}
\lastpage{1}

\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Uncomment next line to change            %%
%% the type of equation numbering           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\numberwithin{equation}{section}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Axiom, Claim, Corollary, Hypothesis, %%
%% Lemma, Theorem, Proposition              %%
%% use \theoremstyle{plain}                 %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{axiom}{Axiom}
\newtheorem{claim}[axiom]{Claim}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Assumption, Definition, Example,     %%
%% Notation, Property, Remark, Fact         %%
%% use \theoremstyle{definition}            %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{example}{Example}
\newtheorem*{fact}{Fact}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Case use \theoremstyle{remark}       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{remark}
\newtheorem{case}{Case}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please put your definitions here:        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\endlocaldefs

\begin{document}

\begin{frontmatter}
\title{Forecasting Influenza Hospitalizations Using a Bayesian Hierarchical Nonlinear Model with Discrepancy}
%\title{A sample article title with some additional note\thanksref{t1}}
\runtitle{Bayesian nonlinear flu hospitalizations forecasting}
%\thankstext{T1}{A sample additional note to the title.}

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only one address is permitted per author. %%
%% Only division, organization and e-mail is %%
%% included in the address.                  %%
%% Additional information can be included in %%
%% the Acknowledgments section if necessary. %%
%% ORCID can be inserted by command:         %%
%% \orcid{0000-0000-0000-0000}               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\snm{Spencer Wadsworth}~\ead[label=e1]{sgw96@iastate.edu}}
% \author[A]{\fnms{Second}~\snm{Author}\ead[label=e2]{second@somewhere.com}\orcid{0000-0000-0000-0000}}
\and
\author[A]{~\snm{Jarad Niemi}\ead[label=e2]{niemi@iastate.edu}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Department of Statistics,
Iowa State University, Ames, IA 50011}

\runauthor{Wadsworth and Niemi}
\end{aug}

\begin{abstract}
The annual influenza outbreak leads to significant public health and economic burdens making it desirable to have prompt and accurate probabilistic forecasts of the disease spread. The United States Centers for Disease Control and Prevention (CDC) hosts annually a national flu forecasting competition which has led to the development of a variety of flu forecast modeling methods. For the first several years of the competition, the target to be forecast was weekly percentage of patients with an influenza-like illness (ILI), but in 2021 the target was changed to weekly hospitalization counts. Reliable state and national hospitalization data has only been available since 2021, but for ILI the data has been available since 2010 and has been successfully forecast for several seasons.
In this manuscript, we introduce a two component modeling framework for forecasting weekly hospitalizations utilizing both hospitalization data and ILI data. The first component is for modeling ILI data using a dynamic nonlinear Bayesian hierarchical model. The second component is for modeling hospitalizations as a function of ILI. For hospitalization forecasts, ILI is first forecasted and then hospitalizations are forecast with ILI forecasts used as a linear or quadratic predictor. In a simulation study, two ILI forecast models, including one similar to the winning model for two seasons of the CDC forecast competition from \cite{osthus2019dynamic} and a nonlinear Bayesian hierarchical model from \cite{ulloa2019} are compared. Also assessed is the usefulness of including a systematic model discrepancy term in the ILI model. Forecasts of state and national hospitalizations for the 2023-24 flu season are made, and different modeling decisions are compared. We found that including a discrepancy component in the ILI model tends to improve forecasts during certain weeks of the year. We also found that other modeling decisions such as the exact nonlinear function to use in the ILI model or the error distribution for hospitalization models may or may not be better than other decisions, depending on the season, location, or week of the forecast.
\end{abstract}

@@

\begin{keyword}
\kwd{Disease outbreak forecasting}
\kwd{Bayesian hierarchical modeling}
\kwd{Probabilistic forecasting}
\kwd{Model discrepancy}
\end{keyword}

\end{frontmatter}

@@

Every year the seasonal influenza outbreak burdens the public health system by infecting millions, causing an influx of primary care visits and hospitalizations and leading to between 290,000 and 650,000 deaths worldwide \cite[]{whoflufact2023}. \cite{molinari2007annual} estimated the United States' annual economic burden from medical costs, loss of income, and deaths to be over \$87 billion. Accurate forecasting of infectious diseases can inform public decision making and ease the burden of an outbreak \cite[]{turtle2021accurate, lutz2019applying}.
There is a growing consensus that disease forecasts should be probabilistic in nature \cite[]{gneiting2014probabilistic, bracher2021evaluating}, and it has been shown that reporting forecast uncertainty along with predictions may lead to better decision making \cite[]{ramos2013probabilistic, joslyn2012uncertainty, winkler1971probabilistic}.

@@

To better inform public decision making regarding the flu epidemic, in 2013 the United States Centers for Disease Control and Prevention (CDC) organized a national flu forecasting competition, also known as FluSight \cite[]{biggerstaff2016results, mathis2024evaluation,cdcfluforecasting2024}. Originally, over a dozen teams of researchers from academic and industry backgrounds participated in FluSight by contributing their own forecast models. Besides the 2020 season --or the flu season spanning the fall of 2020 and the winter of 2021-- FluSight has been operated annually and researchers outside the CDC have been invited to participate. Initially the target data for forecasts was influenza-like illness (ILI) data. ILI is the proportion of patients who meet a healthcare provider and who display flu like symptoms, and ILI data has been available at the state and national level since the 2010 flu season \cite[]{cdcfluview2023,cdc2024fluviewportal}. 
The collaborative ILI forecasting effort has led to a number of modeling developments in flu forecasting \cite[see references therein for more examples]{mcandrew2021adaptively, osthus2021multiscale, osthus2019dynamic, ulloa2019}, and in their paper's introduction, \cite{osthus2019dynamic} categorized the most commonly used flu forecasting models into four classes including mechanistic models based on differential equation compartmental models, agent based models based on population simulation, machine learning/regression models including data driven machine learning and statistical models, and data assimilation models which are constructed by assimilating mechanistic models into a probabilistic framework. An additional forecast model used in FluSight involves the combination of several forecasts into a single ensemble forecast, which has been shown to perform well relative to individual models \cite[]{mcandrew2021adaptively, ray2020ensemble, yamana2016superensemble}.

@@

The contribution of this manuscript is to introduce a two component framework for modeling HHS hospitalization forecasts where hospitalization data and years of ILI data are used to inform forecast models. The first modeling component is a model of ILI data and the second is a model of hospitalization data with ILI as a predictive covariate.
Herein we use ILI models similar to those in \cite{osthus2019dynamic} and \cite{ulloa2019} for ILI forecasting.
The model of \cite{osthus2019dynamic} is a combined data assimilation and statistical regression model which involves a compartmental model in a probabilistic framework. Their model also includes an additional component for capturing a systematic discrepancy between the deterministic part of the model and the actual data, an idea which was first introduced by \cite{kennedy2001bayesian}. The model in \cite{ulloa2019} is a Bayesian hierarchical regression model with an underlying function intended to capture the trajectory of the seasonal ILI data. Herein, we provide a framework under which discrepancy modeling may be used along with a general function modeling ILI data, and we show the effectiveness of including discrepancy modeling during certain periods of the flu season.

@@

In section \ref{sec:data} we review the ILI and hospitalization data provided by the CDC and targeted by FluSight. 
In section \ref{sec:functions} the modeling framework contributed by this manuscript is given. In the same section, functions similar to those used by \cite{osthus2019dynamic} and \cite{ulloa2019} are defined.
 These functions are the susceptible-infectious-recovered (SIR) compartmental model and the asymmetric Gaussian (ASG) function  respectively. Model fitting and implementation are described at the end of the section. Section \ref{sec:simulation2} is a simulation study where four ILI forecast models and their use in forecasting hospitalizations are compared. Commonly used proper scoring rules \cite[]{gneiting2007strictly}, which are also introduced and defined in section \ref{sec:simulation2}, are used for comparing the forecasts. Forecasting of the 2023 flu outbreak along with assessment and comparison under several selected models is performed in section \ref{sec:analysis}, with the analysis being done using the conventions of FluSight. 
 Finally, the manuscript is concluded in section \ref{sec:conclusion} with general observations and some discussion.

 \section{Flu outbreak data} \label{sec:data}
In this section we introduce and define ILI and hospitalization data and evaluate the data visually. ILI and hospitalization data have been the object of forecasting for FluSight with ILI being the target for the first seven seasons and hospitalizations being the target since the 2022 season. Both of these data were collected at the state, territorial, and national level and were reported at least weekly. Overall the data is reported for 53 locations including the 50 US states, the District of Columbia (DC), Puerto Rico (PR), and at the US national level. We will refer to forecast targets throughout this manuscript. A target is the specific horizon, 1, 2, 3, or 4-weeks ahead, for a specific location and week during the season.

\subsection{Influenza-like illness data}
The US Outpatient Influenza-like Illness Surveillance Network (ILINet) collects information on respiratory illness from outpatient visits to health care providers. Over 3,400 outpatient health care providers in all 50 US states, PR, DC, and the US Virgin Islands report each week the total number of outpatient visits along with the number of ILI cases. An ILI case is defined as a ``fever (temperature of $100^{\circ}$F[$37.8^{\circ}$C] or greater) and a cough and/or a sore throat." Prior to the 2021 season, the definition included ``without a known cause other than influenza" \cite[]{cdcfluview2023}. Because other illnesses such as COVID-19, RSV, and the common cold may induce similar respiratory symptoms, ILI may include patients infected with some disease other than influenza. To know whether or not a sick patient is infected with influenza would require a laboratory test.

@@

Figure \ref{fig:us_ili} shows the ILI data at the national level for flu seasons 2010 to 2023. For most seasons there are 52 weeks listed, but for the 2010, 2015, and 2021 seasons there are 53 weeks because there were 53 Sundays during those seasons. 
To better align with the flu behavior, week 1 is set as the first week of August and week 52 or 53 is the last week in July of the following year. For example week 1 of the 2013 season corresponds to the first week of August 2013, and week 52 of the same season corresponds to the last week of July 2014. This convention is used for the remainder of this manuscript.

\begin{figure}
    \centering
    \includegraphics[scale=.5]{Images/us_ili_seasons.png}
    \caption{Percentage of outpatient visits with an influenza-like illness (ILI) in the US for seasons 2010 to 2023. Week 1 is the first week of August of the year the flu season begins and the last week of the season is the last week of July of the following year.}
    \label{fig:us_ili}
\end{figure}

Notable from the plots in figure \ref{fig:us_ili} is the regular trajectory of the ILI. With the exception of season 2020, the ILI begins low at week 1 and increases as the fall and winter progress until the ILI reaches a peak. As spring progresses to summer, the ILI decreases to low values. As \cite{osthus2019dynamic} point out, there is nearly always either a global or local peak at week 22 which typically corresponds to the week between Christmas and New Year's day. 
Whether local or global, the ILI holiday peak is generally expected and thought to be due to widespread holiday travel, school closure, or other unique social behavior   
\cite[]{ewing2017contact, garza2013effect}.
The only seasons when there was not a peak at week 22 were season 2022 where the season peak occurred particularly early and season 2020 which was greatly influenced by the COVID-19 pandemic.

@@

Figure \ref{fig:us_hosp} shows the weekly national hospitalizations for the 2022 and 2023 flu seasons. These plots show similarities to the ILI plots in \ref{fig:us_ili} in that at the early weeks of the season hospitalizations are low, but they increase in the fall to a peak after which they decrease until the flu outbreak ends. For both the 2022 and 2023 seasons, the hospitalizations peaked during the same week as ILI, and in 2023 that peak occurred during the holiday week 22. Additional plots of ILI data and hospitalization
data for select states are included in the supplementary material 
\cite[]{wadsworth2024bas}. These plots show flu outbreak trajectories similar to 
those in figures \ref{fig:us_ili} and \ref{fig:us_hosp} with activity increasing
to a peak and then decreasing.

\begin{figure}
    \centering
    \includegraphics[scale=.5]{Images/us_hospitalizations.png}
    \caption{Weekly flu confirmed hospitalization counts at the national level for 2022 (left) and 2023 (right) flu seasons}
    \label{fig:us_hosp}
\end{figure}

@@

The typical behavior of the ILI data which starts at lower values in the late summer and fall but which increases to a peak, usually in December or January, followed by a decline motivates the use of a nonlinear function which follows of a similar trajectory for modeling ILI. Compartmental models are standard mathematical models used for disease outbreaks. One important compartmental model is the susceptible-infectious-recovered (SIR) model, which is used by some to model ILI data \cite[]{osthus2019dynamic, allen2017primer}. \cite{ulloa2019} chose to use the asymmetric Gaussian (ASG) function to model ILI data. The SIR and ASG models may both be appropriate to describe ILI behavior over the course of a flu season, however there may also be systematic behavior not captured by either, necessitating an additional model component to capture the discrepancy. In the first part of this section, we present an ILI model similar to the model in \cite{osthus2019dynamic}. With some generalization, the model may incorporate any appropriate nonlinear function, though the focus here is on the SIR and ASG functions which are defined.

@@

The SIR compartmental model is a mathematical model used for modeling disease outbreaks and was introduced by \cite{kermack1927contribution}. Since then, compartmental models have become standard for modeling infectious diseases \cite[]{allen2008mathematical}, and many extensions have been made and studied \cite[for example]{simon2020sir, allen2017primer, van2008deterministic}. The SIR mathematical model includes three compartments and assumes that at any time $t>0$ every individual in a closed population belongs to exactly one compartment. The three compartments are susceptible ($S$), infectious ($I$), and recovered ($R$), and their interaction over the course of an outbreak is described by the differential equations in (\ref{eq:sir_diff}).

>>>>

\documentclass{article}

\usepackage{PRIMEarxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[linktocpage=true]{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
% \usepackage{lipsum}
\usepackage{fancyhdr}       % header
\usepackage{graphicx}       % graphics
\graphicspath{{media/}}     % organize your images and other figures under media/ folder

% Packages added by Spencer
\usepackage{amsmath,amsthm,amssymb,scrextend}
\usepackage{appendix}
\AtBeginEnvironment{appendices}{\renewcommand\thesection{\Alph{section}}}
\usepackage{enumerate}
\usepackage{csquotes}
% \usepackage{subcaption}
\usepackage{amssymb}
% \usepackage{bbm}
\usepackage[bb=pazo]{mathalpha}
\usepackage{bm}
\usepackage{newtxmath} %for making indicator function
\usepackage{booktabs} %for the table in the analysis section
% \usepackage{enumerate} %for the short algorithm
\usepackage{subfigure} %for making plots with multiple images
\usepackage{tabularx} %for the spacing in tabular

\pagestyle{empty}

\newcommand\Bbbbone{%
  \ifdefined\mathbbb%
    \mathbbb{1}%
  \else%
    \boldsymbol{\mathbb{1}}%
  \fi}

% \usepackage{xr}
% \externaldocument{supplementary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \usepackage[pdftex]{graphicx}
% Standard, old-style thesis
% \usepackage{isutraditional}   
% \chaptertitle
% Old-style, thesis numbering down to subsubsection
% \alternate
% \usepackage{rotating}
% Bibliography without numbers or labels
\usepackage{natbib}
% Use the following line if you want square brackets and numbering system
%\usepackage[square, numbers]{natbib}

% this package is for getting certain math scripts like the 1 needed for an indicator function
% \usepackage{bbm}

% using this package to make a better looking list
\usepackage{enumerate}
% try this one for a list as well. may be better
% \usepackage[font=itshape]{quoting}

% package for tables

% \renewcommand{\bibfont}{\setstretch{1}\selectfont}
% \setlength{\bibsep}{13.2pt}

\usepackage{chapterbib}
%\renewcommand\bibsection{\section*{\ref name}}
%\renewcommand\bibsection{\section*}
\renewcommand{\bibsection}{\section{References}}

%\bibliographystyle{apa}  % changed from apa, abbrv, unsrtnat
%\includeonly{titletoc,chapter1}
%Optional Package to add PDF bookmarks and hypertext links
% \usepackage[pdftex,hypertexnames=false,linktocpage=true]{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,anchorcolor=blue,citecolor=blue,filecolor=blue,urlcolor=blue,bookmarksnumbered=true,pdfview=FitB}

%\usepackage{titletoc}
% \usepackage{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Packages/commands added by Jarad
\usepackage{xcolor}
\newcommand{\jarad}[1]{{\color{red} Jarad: #1}}
\newcommand{\spencer}[1]{{\color{blue} Spencer: #1}}

%Header
\pagestyle{fancy}
\thispagestyle{empty}
\rhead{ \textit{ }}

% Update your Headers here
\fancyhead[LO]{Bayesian nonlinear flu hospitalizations forecasting}
% \fancyhead[RE]{Firstauthor and Secondauthor} % Firstauthor et al. if more than 2 - must use \documentclass[twoside]{article}

%% Title
\title{Forecasting Influenza Hospitalizations Using a Bayesian Hierarchical Nonlinear Model with Discrepancy
%%%% Cite as
%%%% Update your official citation here when published 
% \thanks{\textit{\underline{Citation}}: 
% \textbf{Authors. Title. Pages.... DOI:000000/11111.}} 
}

\author{
  Spencer Wadsworth \\
  Iowa State University \\
  Ames, Iowa\\
  \texttt{sgw96@iastate.edu} \\
  %% examples of more authors
   \And
  Jarad Niemi \\
  Iowa State University \\
  Ames, Iowa\\
  \texttt{niemi@iastate.edu} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

\begin{document}
\maketitle
\begin{abstract}
The annual influenza outbreak leads to significant public health and economic burdens making it desirable to have prompt and accurate probabilistic forecasts of the disease spread. The United States Centers for Disease Control and Prevention (CDC) hosts annually a national flu forecasting competition which has led to the development of a variety of flu forecast modeling methods. For the first several years of the competition, the target to be forecast was weekly percentage of patients with an influenza-like illness (ILI), but in 2021 the target was changed to weekly hospitalization counts. Reliable state and national hospitalization data has only been available since 2021, but for ILI the data has been available since 2010 and has been successfully forecast for several seasons.
In this manuscript, we introduce a two component modeling framework for forecasting weekly hospitalizations utilizing both hospitalization data and ILI data. The first component is for modeling ILI data using a dynamic nonlinear Bayesian hierarchical model. The second component is for modeling hospitalizations as a function of ILI. For hospitalization forecasts, ILI is first forecasted and then hospitalizations are forecast with ILI forecasts used as a linear or quadratic predictor. In a simulation study, two ILI forecast models, including one similar to the winning model for two seasons of the CDC forecast competition from Osthus et al.
% \cite[]{osthus2019dynamic} 
% 
and a nonlinear Bayesian hierarchical model from Ulloa 
% \cite[]{ulloa2019}
are compared. Also assessed is the usefulness of including a systematic model discrepancy term in the ILI model. Forecasts of state and national hospitalizations for the 2023-24 flu season are made, and different modeling decisions are compared. We found that including a discrepancy component in the ILI model tends to improve forecasts during certain weeks of the year. We also found that other modeling decisions such as the exact nonlinear function to be used in the ILI model or the error distribution for hospitalization models may or may not be better than other decisions, depending on the season, location, or week of the forecast.
\end{abstract}

% \begin{abstract}
% The annual influenza outbreak leads to significant public health and economic burdens making it desirable to have prompt and accurate probabilistic forecasts of the disease spread. The United States Centers for Disease Control and Prevention (CDC) hosts annually a national flu forecasting competition which has led to the development of a variety of flu forecast modeling methods. Beginning in 2013, the target to be forecast was weekly percentage of patients with an influenza-like illness (ILI), but in 2021 the target was changed to weekly hospitalizations. Reliable hospitalization data has only been available since 2021, but ILI data has been available since 2010 and has been successfully forecast for several seasons. In this manuscript, we introduce a two component modeling framework for forecasting hospitalizations utilizing both hospitalization and ILI data. The first component is for modeling ILI data using a nonlinear Bayesian model. The second component is for modeling hospitalizations as a function of ILI. For hospitalization forecasts, ILI is first forecast then hospitalizations are forecast with ILI forecasts used as a predictor. In a simulation study, the hospitalization forecast model is assessed and two previously successful ILI forecast models are compared. Also assessed is the usefulness of including a systematic model discrepancy term in the ILI model. Forecasts of state and national hospitalizations for the 2023-24 flu season are made, and different modeling decisions are compared. We found that including a discrepancy component in the ILI model tends to improve forecasts during certain weeks of the year. We also found that other modeling decisions such as the exact nonlinear function to be used in the ILI model or the error distribution for hospitalization models may or may not be better than other decisions, depending on the season, location, or week of the forecast.
% \end{abstract}

@@

@@

Every year the seasonal influenza outbreak burdens the public health system by infecting millions, causing an influx of primary care visits and hospitalizations and leading to between 290,000 and 650,000 deaths worldwide \cite[]{whoflufact2023}. Molinari et al. \cite[]{molinari2007annual} estimated the United States' annual economic burden from medical costs, loss of income, and deaths to be over \$87 billion. Accurate forecasting of infectious diseases can inform public decision making and ease the burden of an outbreak \cite[]{turtle2021accurate, lutz2019applying}.
There is a growing consensus that disease forecasts should be probabilistic in nature \cite[]{gneiting2014probabilistic, bracher2021evaluating}, and it has been shown that reporting forecast uncertainty along with predictions may lead to better decision making \cite[]{ramos2013probabilistic, joslyn2012uncertainty, winkler1971probabilistic}.

@@

To better inform public decision making regarding the flu epidemic, in 2013 the United States Centers for Disease Control and Prevention (CDC) organized a national flu forecasting competition, also known as FluSight \cite[]{biggerstaff2016results, mathis2024evaluation,cdcfluforecasting2024}. Originally, over a dozen teams of researchers from academic and industry backgrounds participated in FluSight by contributing their own forecast models. Besides the 2020 season --or the flu season spanning the fall of 2020 and the winter of 2021-- FluSight has been operated annually and researchers outside the CDC have been invited to participate. Initially the target data for forecasts was influenza-like illness (ILI) data. ILI is the proportion of patients who meet a healthcare provider and who display flu like symptoms, and ILI data has been available at the state and national level since the 2010 flu season \cite[]{cdcfluview2023,cdc2024fluviewportal}. 
The collaborative ILI forecasting effort has led to a number of modeling developments in flu forecasting \cite[see references therein for more examples]{mcandrew2021adaptively, osthus2021multiscale, osthus2019dynamic, ulloa2019}, and in their paper's introduction, Osthus et al.  \cite[]{osthus2019dynamic} categorized the most commonly used flu forecasting models into four classes including mechanistic models based on differential equation compartmental models, agent based models based on population simulation, machine learning/regression models including data driven machine learning and statistical models, and data assimilation models which are constructed by assimilating mechanistic models into a probabilistic framework. An additional forecast model used in FluSight involves the combination of several forecasts into a single ensemble forecast, which has been shown to perform well relative to individual models \cite[]{mcandrew2021adaptively, ray2020ensemble, yamana2016superensemble}.

@@

The contribution of this manuscript is to introduce a two component framework for modeling HHS hospitalization forecasts where hospitalization data and years of ILI data are used to inform forecast models. The first modeling component is a model of ILI data and the second is a model of hospitalization data with ILI as a predictive covariate.
Herein we use ILI models similar to those in Osthus et al. \cite[]{osthus2019dynamic} and Ulloa \cite[]{ulloa2019} for ILI forecasting.
The model of Osthus et al. \cite[]{osthus2019dynamic} is a combined data assimilation and statistical regression model which involves a compartmental model in a probabilistic framework. Their model also includes an additional component for capturing a systematic discrepancy between the deterministic part of the model and the actual data, an idea which was first introduced by Kennedy and O'Hagan \cite[]{kennedy2001bayesian}. The model in Ulloa \cite[]{ulloa2019} is a Bayesian hierarchical regression model with an underlying function intended to capture the trajectory of the seasonal ILI data. Herein, we provide a framework under which discrepancy modeling may be used along with a general function modeling ILI data, and we show the effectiveness of including discrepancy modeling during certain periods of the flu season.

@@

In section \ref{sec:data} we review the ILI and hospitalization data provided by the CDC and targeted by FluSight. 
In section \ref{sec:functions} the modeling framework contributed by this manuscript is given. In the same section, functions similar to those used by Osthus et al. \cite[]{osthus2019dynamic} and Ulloa \cite[]{ulloa2019} are defined.
 These functions are the susceptible-infectious-recovered (SIR) compartmental model and the asymmetric Gaussian (ASG) function  respectively. Model fitting and implementation are described at the end of the section. Section \ref{sec:simulation2} includes a simulation study where four ILI forecast models and their use in forecasting hospitalizations are compared. Commonly used proper scoring rules \cite[]{gneiting2007strictly}, which are also introduced and defined in section \ref{sec:simulation2}, are used for comparing the forecasts. Forecasting of the 2023 flu outbreak along with assessment and comparison under several selected models is performed in section \ref{sec:analysis}. 
 Finally, the manuscript is concluded in section \ref{sec:conclusion} with general observations and some discussion.

 \section{Flu outbreak data} \label{sec:data}
In this section we introduce and define ILI and hospitalization data and visually evaluate the data. ILI and hospitalization data have been the object of forecasting for FluSight with ILI being the target for the first seven seasons and hospitalizations being the target since the 2022 season. Both of these data were collected at the state, territorial, and national level and were reported at least weekly. Overall the data is reported for 53 locations including the 50 US states, the District of Columbia (DC), Puerto Rico (PR), and at the US national level. We will refer to forecast targets throughout this manuscript. A target is the specific horizon, 1, 2, 3, or 4-weeks ahead, for a specific location and week during the season.

\subsection{Influenza-like illness data}
The US Outpatient Influenza-like Illness Surveillance Network (ILINet) collects information on respiratory illness from outpatient visits to health care providers. Over 3,400 outpatient health care providers in all 50 US states, PR, DC, and the US Virgin Islands report each week the total number of outpatient visits along with the number of ILI cases. An ILI case is defined as a ``fever (temperature of $100^{\circ}$F[$37.8^{\circ}$C] or greater) and a cough and/or a sore throat." Prior to the 2021 season, the definition included ``without a known cause other than influenza" \cite[]{cdcfluview2023}. Because other illnesses such as COVID-19, RSV, and the common cold may induce similar respiratory symptoms, ILI may include patients infected with some disease other than influenza. To know whether or not a sick patient is infected with influenza requires a laboratory test.

@@

The top of figure \ref{fig:us_ili} shows the ILI data at the national level for flu seasons 2010 to 2023. For most seasons there are 52 weeks listed, but for the 2010, 2015, and 2021 seasons there are 53 weeks because there were 53 Sundays during those seasons. 
To better align with the flu behavior, week 1 is set as the first week of August and week 52 or 53 is the last week in July of the following year. For example week 1 of the 2013 season corresponds to the first week of August 2013, and week 52 of the same season corresponds to the last week of July 2014. This convention is used for the remainder of this manuscript.

\begin{figure}[hbt!]
    \centering
    \includegraphics[scale=.5]{Images/us_ili_seasons.png}

    \centering
    \includegraphics[scale=.4]{Images/us_hospitalizations.png}
    \caption{Percentage of outpatient visits with an influenza-like illness (ILI) in the US for seasons 2010 to 2023 (top) and weekly flu confirmed hospitalization counts at the national level for 2022 and 2023 flu seasons (bottom). Week 1 is the first week of August of the year the flu season begins and the last week of the season is the last week of July of the following year.}
    \label{fig:us_ili}
\end{figure}

Notable from the plots in the top of figure \ref{fig:us_ili} is the regular trajectory of the ILI. With the exception of season 2020, the ILI begins low at week 1 and increases as the fall and winter progress until the ILI reaches a peak. As spring progresses to summer, the ILI decreases to low values. As Osthus et al. \cite[]{osthus2019dynamic} point out, there is nearly always either a global or local peak at week 22 which typically corresponds to the week between Christmas and New Year's day. 
Whether local or global, the ILI holiday peak is generally expected and thought to be due to widespread holiday travel, school closure, or other unique social behavior   
\cite[]{ewing2017contact, garza2013effect}.
The only seasons when there was not a peak at week 22 were season 2022 where the season peak occurred particularly early and season 2020 which was greatly influenced by the COVID-19 pandemic.

@@

The bottom of figure \ref{fig:us_ili} shows the weekly national hospitalizations for the 2022 and 2023 flu seasons. These plots show similarities to the ILI plots in the top of the figure in that at the early weeks of the season hospitalizations are low, but they increase in the fall to a peak after which they decrease until the flu outbreak ends. For both the 2022 and 2023 seasons, the hospitalizations peaked during the same week as ILI, and in 2023 that peak occurred during the holiday week 22. Additional plots of ILI data and hospitalization
data for select states are included in the supplementary material 
\cite[]{wadsworth2024bas}. These plots show flu outbreak trajectories similar to 
those in figure \ref{fig:us_ili} with activity increasing
to a peak and then decreasing.

% \begin{figure}
%     \centering
%     \includegraphics[scale=.5]{Images/us_hospitalizations.png}
%     \caption{Weekly flu confirmed hospitalization counts at the national level for 2022 (left) and 2023 (right) flu seasons}
%     \label{fig:us_hosp}
% \end{figure}

@@

The typical behavior of the ILI data which starts at lower values in the late summer and fall but which increases to a peak, usually in December or January, followed by a decline motivates the use of a nonlinear function which follows a similar trajectory. Compartmental models are standard mathematical models used for modelling disease outbreaks. One important compartmental model is the susceptible-infectious-recovered (SIR) model, which is used by some to model ILI data \cite[]{osthus2019dynamic, allen2017primer}. Ulloa \cite[]{ulloa2019} chose to use the asymmetric Gaussian (ASG) function to model ILI data. The SIR and ASG models may both be appropriate to describe ILI behavior over the course of a flu season, however there may also be systematic behavior not captured by either, necessitating an additional model component to capture the discrepancy. In the first part of this section, we present an ILI model similar to the model in Osthus et al. \cite[]{osthus2019dynamic}. With some generalization, the model may incorporate any appropriate nonlinear function, though the focus here is on the SIR and ASG functions which are defined.

@@

The SIR compartmental model is a mathematical model used for modeling disease outbreaks and was introduced by Kermack and McKendrick \cite[]{kermack1927contribution}. Since then, compartmental models have become standard for modeling infectious diseases \cite[]{allen2008mathematical}, and many extensions have been made and studied \cite[for example]{simon2020sir, allen2017primer, van2008deterministic}. The SIR mathematical model includes three compartments and assumes that at any time $t>0$ every individual in a closed population belongs to exactly one compartment. The three compartments are susceptible ($S$), infectious ($I$), and recovered ($R$), and their interaction over the course of an outbreak is described by the differential equations in (\ref{eq:sir_diff}).

\begin{equation}
    \label{eq:sir_diff}
    \frac{dS}{dt} = -\beta SI, \quad \frac{dI}{dt} = \beta S I - \delta I, \quad \frac{dR}{dt} = \delta I
\end{equation}
Here $S$, $I$, and $R$ represent the proportion of the population in each compartment such that $S + I + R = 1$ for all $t$. The trajectory is determined by the disease transmission rate $\beta > 0$ and the recovery rate $\delta > 0$. Respectively, these may be thought of as the expected proportion of susceptible individuals who will be infected by an infectious individual and the expected rate of recovery to an immune state for a newly infected person. Whether or not a disease outbreak is classified as an epidemic is determined by the initial susceptible population $S_0$, or the susceptible population at time $0$, and the parameter $\rho = \delta/
\beta$. If $S_0/\rho > 1$, the outbreak is considered an epidemic. It is non-epidemic if $S_0/\rho \leq 1$ \cite[]{osthus2019dynamic}. Figure \ref{fig:sir_traj} shows the trajectory of the three compartments of an SIR model where $S_0$ and $\rho$ were selected to match an outbreak that is an epidemic. In the case where $S_0 \leq \rho$, the trajectory for the $I$ compartment would never be increasing. The increase to a peak and subsequent decrease in the $I$ compartment of figure \ref{fig:sir_traj} suggest it is reasonable to model ILI by this compartment. Thus in modeling the ILI data, we consider the data to be analogous to the $I$ proportion of the population.
